Work Fragmentation in Developer Interaction Data

### 2015-08-07 paper.tex
intro & related

\begin{abstract}
%\boldmath
Information workers and software developers are exposed to work fragmentation, an interleaving of activities and interruptions during their normal work day. Small-scale observational studies have shown that this can be detrimental to their work. In this paper, we perform a large-scale study of this phenomenon for the particular case of software developers performing software evolution tasks. Our study is based on several thousands interaction traces collected by Mylyn and Usage Data Collector. We observe that work fragmentation is correlated to lower observed productivity at both the macro level (for entire sessions), and at the micro level (around markers of work fragmentation); further, longer activity switches seem to strengthen the effect. These observations are basis for subsequent studies investigating the phenomenon of work fragmentation.
\end{abstract}

@@

In this paper, we present a study addressing these issues by (1) being focused on software developers, and (2) using MSR techniques to base its conclusions on a much larger amount of data (specifically, thousands of development sessions recorded by Mylyn). The downside is that the data exploited, being operational data \cite{M14}, is of lower quality than data extracted from a full-fledged observational study. Trading quality for quantity allows us to explore different aspects of the problem, and to generate hypotheses for subsequent studies of work fragmentation and interruptions in software development. We address the following three questions:
\begin{itemize}
\item \textbf{RQ1:} What is the relationship between the observed interruptions and the observed developer productivity? 
\item \textbf{RQ2:} Is the observed relationship more pronounced in the presence of prolonged interruptions?
\item \textbf{RQ3:}  What is the observed relationship in the vicinity of interruptions?
\end{itemize}

Additionally, the Eclipse Usage Data Collector (UDC) interaction data allows us to dig further into more specific subjects resulting in the following complementary research questions:
\begin{itemize}
\item \textbf{RQ4:} How is the developer's performance during the recovery time after an interruption? 
\item \textbf{RQ4:} What events are more common during recovery time?
\end{itemize}

\emph{A note on terminology}: in this paper we (ab)use the word \emph{interruption} to refer to a \emph{gap of activity} observed in IDE activity. We do not presume that all of these activity gaps are actual interruptions (our results lead us to believe otherwise); rather, we assume they are indicators of work fragmentation in general (that is, a programmer switching contexts to perform other activities while developing, such as answering email, looking for the solution of a problem on the Internet, etc). Of note, work by Zou and Godfrey \cite{ZG06} found that interruptions were a common cause for a lack of observed activity, and classified periods of inactivity as interruptions. %programmer feedback

\textbf{Structure of the paper.} We start with a literature review of studies of work fragmentation in Section 2. This section also contains a review of studies performed with the Mylyn dataset that we use. In Section 3, we describe the Mylyn dataset in further detail, highlighting potential issues in the operational data that we use in the analysis. In Section 4, we describe the measurements used in this study, and detail the processing of the data that is applied to alleviate the issues described in Section 3. The same section shows how we convert interaction traces to multivariate time series representing development sessions, and present the metrics in the analysis. In Section 5, we answer our first research question: we find an inverse relationship between number of interruptions and our three productivity indicators. In Section 6, we answer our second research question, finding that this relationship is more pronounced for development sessions with at least one longer interruption. In Section 7 we answer our third research question and find that our productivity indicators are indeed lower in the vicinity of interruptions; we further find three patterns of interruptions with different characteristics. In Section 8 we describe the Usage Data Collector dataset and give detail on the selection and transformation procedures. In Section 9, we compare the Mylyn data with Usage Data Collector by executing the same processes to obtain the above results and we find that all of the previous conclusions agree with this new data. In Section 10 we answer our fourth research question, finding a positive evolution of productivity during the recovery time and a relation with the effect of the interruption. In Section 11, we answer our fifth research question finding differences on the activities performed after a positive or negative interruption, judging by the type of events executed. We close the paper by discussing the threats to validity of this study in Section 12, before concluding in Section 13.

@@

\subsection{Empirical Studies with Usage Data Collector}
The Usage Data Collector dataset is a large collection of information about interaction data of Eclipse's users. It is publicly available and has been used on many subjects around mining interaction data. On refactoring activities, Murphy-Hll et al. used this data to understand how programmers are using refactoring tools analyzing patterns on refactoring practices, finding that most programmers don't make deep use of this tools, leaving untouched most of the configuration parameters and performing manually most of the refactoring  \cite{MPB12}. Liu et al. identified refactoring tactics of the type data floss and root canal, finding that the majority of them are floss refactoring and detecting differences on activities performed depending on the type of the refactoring \cite{LGN12}. Sahin et al. analyzed the energy usage changes after refactoring and used the UDC data to identify the most common refactoring events and selected those that fulfill the requirements of being common and cause structural changes \cite{SPC14}.

On coding tools and strategies, Murphy-Hill et al. investigated on command recommendation for software developers depending on past history of command usage and tested a set of algorithms using UDC data to obtain CVS and editing commands, getting positive feedback on a live evaluation with users \cite{MJM12}. Yoon et al. uses UDC data to compare the results obtained from a proposed plug-in for the capture of low-level events, finding a very similar report on command distribution from both tools \cite{YM11}.

>>>>

\begin{abstract}
%\boldmath
Information workers and software developers are exposed to work fragmentation, an interleaving of activities and interruptions during their normal work day. Small-scale observational studies have shown that this can be detrimental to their work. In this paper, we perform a large-scale study of this phenomenon for the particular case of software developers performing software evolution tasks. Our study is based on several thousands interaction traces collected by Mylyn and the Eclipse Usage Data Collector. We observe that work fragmentation is correlated to lower observed productivity at both the macro level (for entire sessions), and at the micro level (around markers of work fragmentation); further, longer activity switches seem to strengthen the effect. These observations are basis for subsequent studies investigating the phenomenon of work fragmentation.
\end{abstract}

@@

In this paper, we present a study addressing these issues by (1) being focused on software developers, and (2) using MSR techniques to base its conclusions on a much larger amount of data (specifically, thousands of development sessions recorded by Mylyn and the Eclipse Usage Data Collector). The downside is that the data exploited, being operational data \cite{M14}, is of lower quality than data extracted from a full-fledged observational study. Trading quality for quantity allows us to explore different aspects of the problem, and to generate hypotheses for subsequent studies of work fragmentation and interruptions in software development. We address the following three questions:
\begin{itemize}
\item \textbf{RQ1:} What is the relationship between the observed interruptions and the observed developer productivity? 
\item \textbf{RQ2:} Is the observed relationship more pronounced in the presence of prolonged interruptions?
\item \textbf{RQ3:}  What is the observed relationship in the vicinity of interruptions?
\end{itemize}

Additionally, the Eclipse Usage Data Collector (UDC) interaction data allows us to dig further into more specific aspects, resulting in the following complementary research questions:
\begin{itemize}
\item \textbf{RQ4:} How is the developer's performance during the recovery time after an interruption? 
\item \textbf{RQ5:} What events are more common during recovery time?
\end{itemize}

\emph{A note on terminology}: in this paper we (ab)use the word \emph{interruption} to refer to a \emph{gap of activity} observed in IDE activity. We do not presume that all of these activity gaps are actual interruptions (our results lead us to believe otherwise); rather, we assume they are indicators of work fragmentation in general (that is, a programmer switching contexts to perform other activities while developing, such as answering email, looking for the solution of a problem on the Internet, etc). Of note, work by Zou and Godfrey \cite{ZG06} found that interruptions were a common cause for a lack of observed activity, and classified all periods of inactivity as interruptions. %programmer feedback

\textbf{Structure of the paper.} We start with a literature review of studies of work fragmentation in Section 2. This section also contains a review of studies performed with the Mylyn and UDC datasets that we use. In Section 3, we describe the Mylyn dataset in further detail, highlighting potential issues in the operational data that we use in the analysis. Section 4, describes the measurements used in this study, and details the processing of the data that is applied to alleviate the issues described in Section 3. The same section shows how we convert interaction traces to multivariate time series representing development sessions, and present the metrics in the analysis. In Section 5, we answer our first research question: we find an inverse relationship between number of interruptions and our three productivity indicators. In Section 6, we answer our second research question, finding that this relationship is more pronounced for development sessions with at least one longer interruption. In Section 7 we answer our third research question and find that our productivity indicators are indeed lower in the vicinity of interruptions; we further find three patterns of interruptions with different characteristics. Starting in section 8, we focus on the UDC dataset; this section describes the dataset and gives details on the selection and transformation procedures. In Section 9, we replicate our first three research questions with the UDC data; we find that all of the previous conclusions agree with this new data, although the size of the observed effects differ. Section 10 answers our fourth research question, finding a positive evolution of productivity during the recovery time, and a relation with the effect of the interruption. Finally Section 11 answers our fifth research question; based on the type of events executed, we find differences in the activities performed after a positive or negative interruption. We close the paper by discussing the threats to validity of this study in Section 12, before concluding in Section 13.

@@

\subsection{Empirical Studies with Usage Data Collector}
The Usage Data Collector dataset is a large collection of information about interaction data of Eclipse's users. It is publicly available and has been used in several studies mining interaction data.

On refactoring activities, Murphy-Hll et al. used this data to understand how programmers are using refactoring tools analyzing patterns on refactoring practices, finding that most programmers don't make deep use of these tools, leaving untouched most of the configuration parameters and performing manually most of the refactoring  \cite{MPB12}. Liu et al. identified refactoring tactics (floss refactoring and root canal refactoring, following Murphy-Hill's terminology), finding that the majority of them are floss refactoring and detecting differences on activities performed depending on the type of the refactoring \cite{LGN12}. Sahin et al. analyzed the energy usage changes after refactoring and used the UDC data to identify the most common refactoring events and selected those that fulfill the requirements of being common and cause structural changes \cite{SPC14}.

On coding tools and strategies, Murphy-Hill et al. investigated command recommendations for software developers depending on their past history of command usage and tested a set of algorithms using UDC data to obtain CVS and editing commands, getting positive feedback on a live evaluation with users \cite{MJM12}. Yoon et al. use UDC data to compare the results obtained from a proposed plug-in for the capture of low-level events, finding a very similar report on command distribution from both tools \cite{YM11}.
